---
title: "NeDCwor"
author: "MohammedBaaoum"
date: "11/24/2020"
output: html_document
---

```{r}
library(MASS)
library(car)
library(pROC)
library(rpart)
library(boot)
library(QuantPsyc)
library(lm.beta)
```

•	Patient  Length of Stay (LOS)  is an important indicator of emergency department (ED) performance that correlates strongly with overcrowding 
•	Prolonged  LOS is viewed as evidence of poor hospital performance and has been associated with a higher mortality rate  and lower patient satisfaction (Pitts et al., 2014, Singer, 2011).
•	It is estimated that reducing ED LOS by 1 hour could decrease the number of deaths in high-risk  patients by 6.5% and by almost 13% in lower-risk patients (Barish, & Arnold  , 2012). 


# Objective: Examine the association between LOS and multi-level factors including patient-, service- and organization-level factors. 
#	RQ1.  What patient-related factors have a significant association with patient length of stay? 
#	RQ2.  What service-related factors have a significant association with patient length of stay? 
#	RQ3.  What organizational factors have a significant association with patient length of stay? 




#Data filted2 after delting patient who went to Norht Phd after entinring normal posds
```{r}
EDdata3 <- read.csv ("SelctedDataFilter2New_MergAmb.csv", header=TRUE)
names(EDdata3)
```
# Select relvant variables  based on stakholder intreve and lit review 

#Patient Factors: Gender,Age.Group,  ESILevel, TransportArrive, Arrival.Shift..1st..12.AM.,FinalDiagnosis,DispoType,Month, 

# service :  Ward.Change, FirstWard, EDLevel

# Organizational: PatientsPerMD_Mean_Rd,PatientsPerNURSE_Mean, PatientsPerPA_RESIDENT_Mean_Rd , 
```{r}
LOSnoNorthTrans <- EDdata3$LOS.rounded
EDdatanoNorthTrans <-cbind(LOSnoNorthTrans,EDdata3[,c(5,9,21,22,24,25,27,30,31,34,38,42,44,46)])
names(EDdatanoNorthTrans)
```


#Check LOS 
```{r}
require(lattice)
require(openintro)
histogram(x= EDdatanoNorthTrans$LOSnoNorthTrans,breaks = 20, data=EDdatanoNorth)
densityplot(x= EDdatanoNorthTrans$LOSnoNorthTrans,data=EDdatanoNorth)
```

# Charactrestic of patient byond 24 hours in ED
```{r}
EDdataByond24_No_NorthTrans<-subset(EDdatanoNorthTrans,EDdatanoNorthTrans$LOSnoNorthTrans>24)
EDdataByond24_No_North_Trans_noNA<- na.omit(EDdataByond24_No_NorthTrans)  #EDdata4 same as ED3.5 but we removed NA 
table(EDdataByond24_No_North_Trans_noNA$FinalDiagnosis)
```
#44 observation majoirity are metnal ilneessed . As disscussed wit stakholders they considred outlier and decided not to include them in the model. 

___________________Delting Values More than 24 hours and Delting NA values-------


#Data after eliminating North & Sounth, Beharioal probelm and Transfer to North the outliers, Elimating data byond 24 hours ( Data contain NA First wards)
```{r}
EDdataBelow24_No_NorthTrans<-subset(EDdatanoNorthTrans,EDdatanoNorthTrans$LOSnoNorthTrans<24)
EDdataBelow24_No_NorthTrans2<- na.omit (EDdataBelow24_No_NorthTrans) ####Delet NA Values (560 Value NA)
tail(EDdataBelow24_No_NorthTrans2)
```

#LOS histogram 
```{r}
require(lattice)
require(openintro)
histogram(x= EDdataBelow24_No_NorthTrans2$LOSnoNorthTrans,breaks = 20, data=EDdataBelow24_No_NorthTrans2)
densityplot(x= EDdataBelow24_No_NorthTrans2$LOSnoNorthTrans,data=EDdataBelow24_No_NorthTrans2)

```


# More mental iliness casses
#Removing Suicidal ideation and Depression and Psychosis (43), Observation 
#Check how it goes if I elminate all the metal inless patien from the data. 

```{r}
EDdataBelow24_No_NorthTrans2_NoSI <- subset(EDdataBelow24_No_NorthTrans2,EDdataBelow24_No_NorthTrans2$FinalDiagnosis != "Suicidal ideation") #deliting socidal ideadtion

EDdataBelow24_No_NorthTrans2_NoSIORDep <- subset(EDdataBelow24_No_NorthTrans2_NoSI ,EDdataBelow24_No_NorthTrans2_NoSI$FinalDiagnosis != "Depression") ## No SI Nor Depression 

EDdataBelow24_No_NorthTrans2_NOmental <-  subset(EDdataBelow24_No_NorthTrans2_NoSIORDep ,EDdataBelow24_No_NorthTrans2_NoSIORDep$FinalDiagnosis != "Psychosis")
```


#LOS histogram after elminating SI and Depresstion 
```{r}
require(lattice)
require(openintro)
histogram(x= EDdataBelow24_No_NorthTrans2_NOmental$LOSnoNorthTrans,breaks = 20, data=EDdataBelow24_No_NorthTrans2_NOmental)
densityplot(x= EDdataBelow24_No_NorthTrans2_NOmental$LOSnoNorthTrans,data=EDdataBelow24_No_NorthTrans2_NOmental)

```

# rename and orgnize the dataset

# Data withouht Noth or Sounth(First or Last), no Mental inlness(SI, Depression and Psycosis), no NA, no NA or Negative valeues
````{r}

LOSfselcted <- EDdataBelow24_No_NorthTrans2_NOmental$LOSnoNorthTrans
EDdatfselcted <-cbind(LOSfselcted,EDdataBelow24_No_NorthTrans2_NOmental[,c(2:15)])
names(EDdatfselcted)

write.csv(EDdatfselcted, file = "EDdatfselcted.csv")

```


# converting Patient Per MD and Patient Per Nurse to Numric to calculate corr
```{r}
EDdatfselcted$PatientsPerMD_Mean_Rd<- as.numeric(as.character(EDdatfselcted$PatientsPerMD_Mean_Rd))
EDdatfselcted$PatientsPerNURSE_Mean_Rd<- as.numeric(as.character(EDdatfselcted$PatientsPerNURSE_Mean_Rd))
EDdatfselcted$PatientsPerPA_RESIDENT_Mean_Rd <- 
as.numeric(as.character(EDdatfselcted$PatientsPerPA_RESIDENT_Mean_Rd))

na.omit(EDdatfselcted)


```



#test correlathion between numerical valeus 

Teset Correlathions
```{r}

Org <- cbind(EDdatfselcted[,c(11,12,13)])

# Combine the columns into a matrix
#Org <- do.call(cbind, Org)
# Calculate the correlation matrix
cor_matrix <- cor(Org, use = "pairwise.complete.obs")
cor_matrix 
```


# Ploting to test linearity assumption 

```{r}
par(mfrow = c(2, 2), mai = c(0.6, 0.5, 0.2, 0.1))
for (i in 2:15) {
  x <- EDdatfselcted[, i]
  y <- EDdatfselcted$LOSfselcted
  valid_indices <- complete.cases(x, y) & is.finite(x) & is.finite(y)
  
  if (any(valid_indices)) {
    plot(x[valid_indices], y[valid_indices], xlab = names(EDdatfselcted)[i], ylab = "Length of Stay", col = 3, pch = 20)
  }
}


```


#Transfomation (Log) before regression to reslove noninearity issues, rename variables
```{r}
#names(EDdatfselcted)

EDdataTranform<- data.frame(LOS= log(EDdatfselcted$LOSfselcted), Arrival_Shift = EDdatfselcted$Arrival.Shift..1st..12.AM. ,Arrival_Day = EDdatfselcted$Arrival.Day, ESLIndex = EDdatfselcted$ESILevel, gender= EDdatfselcted$Gender,Age_group= EDdatfselcted$Age.Group, EDlevel= EDdatfselcted$EDLevel ,Discharge_type = EDdatfselcted$DispoType, Arrival_Mode= EDdatfselcted$TransportArrive, PA_MD= log(EDdatfselcted$PatientsPerMD_Mean_Rd), PA_NA=log(EDdatfselcted$PatientsPerNURSE_Mean_Rd) , PA_Res= log(EDdatfselcted$PatientsPerPA_RESIDENT_Mean_Rd) , first_ward= EDdatfselcted$FirstWard, Ward_Trans= EDdatfselcted$Ward.Change)


names(EDdataTranform)
```

## Taking off the week end (limt the study to week days) 
subset(EDdatanoNorthTrans,EDdatanoNorthTrans$LOSnoNorthTrans<24)                     

```{r}
EDWeekdays <- subset(EDdataTranform,EDdataTranform$Arrival_Day !="Sun")
EDWeekdays1 <- subset(EDWeekdays,EDWeekdays$Arrival_Day!= "Sat")

```

# eliminate direct admit observation 

```{r}
NoDirectadmit <-subset(EDWeekdays1,EDWeekdays1$ESLIndex!= "Direct Admit")
```


# Ensure no NA before modeling 
```{r}
NoDirectadmitNoNa<- na.omit(NoDirectadmit)
```

----------------------------------Model Bulding-----------------
# Step by step  model bulding ( patient charactrisitcs)

```{r}

model1org<- lm(LOS~ gender+ Age_group+ESLIndex+Arrival_Shift+Arrival_Mode+Discharge_type, data =EDdataTranform)
summary(model1org)

```


# Model2--adding service related variabels : first_ward + Ward_Trans+ EDlevel
```{r}
model2org <- lm(LOS~ gender+ Age_group +ESLIndex+Arrival_Shift+Arrival_Mode+Discharge_type+first_ward+Ward_Trans+EDlevel, data =NoDirectadmit )

summary(model2org)

```


## Add orgnizational variable 

##Model3 : PA_NA+PA_MD+ PA_Res
```{r}
install.packages("car")
library(car)
library(lm.beta)
model3org <- lm(LOS~ gender+ Age_group +ESLIndex+Arrival_Shift+Arrival_Mode+Discharge_type+first_ward+Ward_Trans+EDlevel+ PA_NA+PA_MD+ PA_Res, data =NoDirectadmit )
summary(model3org)
vif(model3org)


```

# Residual for Model 3 with tranfroming LOS 
```{r}
par(mfrow=c(2,2), mai=c(.6,.5,.2,.1))
plot(model3org$fitted, model3org$resid, col=2, pch=20)
abline(h=0, col=4, lty=2)
qqnorm(rstudent(model3org), col=2, pch=20, main="")
abline(a=0,b=1,col=4)
hist(rstudent(model3org), col=2, xlab="Studentized Residuals", main="")
```


